---
title: "How to Get More Out of R"
author: "Paul Stevenson"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  ioslides_presentation:
    template: assets/templates/ioslides.html
    logo: assets/images/logo800.jpg
    css: assets/css/ioslides.css
    widescreen: true
    incremental: true
vignette: >
  %\VignetteIndexEntry{How to Get More Out of R}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
  
---

```{r init, include = FALSE, echo = FALSE}
library(biometrics)
library(lubridate)
library(tidyverse)
library(kableExtra)
```

## Session resources

[https://github.com/TelethonKids/RWorkshop](https://github.com/TelethonKids/RWorkshop)

[Worked example](https://github.com/TelethonKids/RWorkshop/tree/master/R/03_data_manipulation)
[Slides](https://github.com/TelethonKids/RWorkshop/tree/master/inst/doc/reproducable-research.html)

## How to Get More Out of R (FD 2h) - PS

- Style Guide
- Mapping functions over data
- Multi-core processing
- R Markdown

# Style Guide

## Why Bother?

<blockquote class="blockquote">
    Good coding style is like correct punctuation: you can manage without it, butitsuremakesthingseasiertoread.
    <footer class="blockquote-footer">Hadley Wickham in <cite title="Source Title">The tidyverse style guide</cite></footer>
</blockquote>

- Well structured and documented code makes it easier for you to find possible mistakes
- It will give your script some structure and a logical flow
- Do your future self a favour!

## Object Names

- Name your variables in a human friendly way
    * e.g. `24_week_blood_pressure` vs. `bp24w`
- snake_case vs CamelCase
    * Be consistent in whatever you choose
    * My preference is snake_case
    * all lower caps with words separated by an underscore "_"

## Comments

- You don't need to comment every line of code
- Place a comment for each "block" of code describing it's purpose
- Single line comments are created with a "#"

```{r}
# This is commented code

```

```{r}
#' R Studio will automatically follow a line starting with "#'" with a new comment

```

- If the comment ends with four "-" or "#" then the following code is collapsable

```{r}
# Example ----

```

# Mapping functions over data

- `?map`
- Functions can be applied to data on a row-by-row basis with the `purrr` function
    * `map` - 
    * `pmap` - 
    * `map2` - 
- Map functions always return a list, these can be changed to return vectors of objects with the same type
    * `map_lgl()`, `map_int()`, `map_dbl()` and `map_chr()`
- `map_dfr()` and `map_dfc()` return data frames created by row-binding and column-binding, respectively
- These can be easily modified to run over multiple processor cords with the `furrr` function


## Overview of base v tidyverse [history/evolution]

# Worked example

## Files

Data is available from GitHub repository ......

## Read in raw data 

    library(readr)
    
    read_csv("/path/to/file.csv", options ...)

## <code>read_csv()</code> options

```{r read_csv_options, echo = F}
tibble(
  Option = c("col_names", "na", "trim_ws", "skip", "n_max", "guess_max", "skip_empty_rows"),
  Description = c("If TRUE, the first row of the input will be used as the column names, and will not be included in the data frame.",
                  "Character vector of strings to interpret as missing values.",
                  "Should leading and trailing whitespace be trimmed from each field before parsing it?",
                  "Number of lines to skip before reading data.",
                  "Maximum number of records to read.",
                  "Maximum number of records to use for guessing column types.",
                  "Should blank rows be ignored altogether?")) %>%
  mutate(Option = paste0("<div style = 'font-size: 0.5em;'><code>", Option, "</code></div>"),
         Description = paste0("<div style = 'font-size: 0.5em;'>", Description, "</div>")) %>% # add <code> tag to all options
  kable("html", escape = F) %>%
  kable_styling("hover", full_width = F)

```

More help can be found by typing `?read_delim` into the R console.


## Other ways to read in data

    load("path/to/file.RData")
    
    library(readr)
    read_rds("path/to/file.rds")

    library(readxl)
    read_xlxs("path/to/file.xlsx")

    library(readstata13)
    readstata13("path/to/file.sta")
    
    Foreign library (SAS, SPSS, ...)
    
Data can also be read directly from a database into R!

## Data frame (tibble)

A data frame is a table (or 2 dimensional array-like object) where each column
stores the data of one variable and each row represents a single record.

These are the properties of a data frame:

* All columns should have a unique name
* All data in one column will be the same class, however each column can be different
* Each column should have the same number of data items/rows
* Rows are not named

Vectors of data are stored in columns:

    c("A", "B", "C", "D")

## Expand on classes (data/time [lubridate - year], string processing)

* Numeric
* Character
* Logical
* Factor
* Date

* Missing data - NA (there is only one missing data code in R)

# Summarise/tidy (cleaning)

```{r data}
load("../data/dat.RData")

```

```{r str_data}
str(tki_demo)

```

## Data summary - high level descriptives

```{r summary_data}
summary(tki_demo)

```

## Looking at the data

```{r head_data}
head(tki_demo)

```

## Create columns/conditional create (if/else)

New column can be created with `mutate()`

```{r ifelse}
tki_demo %>%
  mutate(age = interval(dob, Sys.Date()) %>% as.duration() %>% as.numeric("years") %>% round(1),
         teenager = ifelse(age >= 13, T, F)) %>%
  select(id, dob, age, teenager) %>%
  head()

```

## <code>case_when()</code>

```{r case_when, eval = F}
tki_demo %>%
  mutate(age = interval(dob, Sys.Date()) %>%
           as.duration() %>%
           as.numeric("years") %>%
           round(1),
         age_category = case_when(
           age < 12 ~ "Younger than 12 years",
           age < 14 ~ "12 - 13 years",
           age < 16 ~ "14 - 15 years",
           age < 18 ~ "15 - 17 years",
           T ~ "Older than 18 years"
         ),
         factor(age_category,
                levels = c("Younger than 12 years","12 - 13 years",
                           "14 - 15 years", "15 - 17 years",
                           "Older than 18 years"))) %>%
  select(id, dob, age, age_category) %>%
  head()

```

## <code>case_when()</code>

```{r case_when_output, echo = F}
tki_demo %>%
  mutate(age = interval(dob, Sys.Date()) %>%
           as.duration() %>%
           as.numeric("years") %>%
           round(1),
         age_category = case_when(
           age < 12 ~ "Younger than 12 years",
           age < 14 ~ "12 - 13 years",
           age < 16 ~ "14 - 15 years",
           age < 18 ~ "15 - 17 years",
           T ~ "Older than 18 years"
         ),
         factor(age_category,
                levels = c("Younger than 12 years","12 - 13 years",
                           "14 - 15 years", "15 - 17 years",
                           "Older than 18 years"))) %>%
  select(id, dob, age, age_category) %>%
  head()

```

## Parsing dates with <code>lubridate</code>

* Run some examples - adding day if YYYYMM data provided
* Converting char to date
* example with date and time

# Subset/merge/join

## needs fixing

So far all the tools you’ve learned have worked with complete data frames. If you want to pull out a single variable, you need some new tools, $ and [[. [[ can extract by name or position; $ only extracts by name but is a little less typing.

df <- tibble(
  x = runif(5),
  y = rnorm(5)
)

## Extract by name
df$x

df[["x"]]


## Extract by position
df[[1]]

To use these in a pipe, you’ll need to use the special placeholder .:

df %>% .$x

df %>% .[["x"]]

# Gather/spread/melt/cast/separate/reshape

## Long/wide

Gather/spread

## Functions

Don't repeat yourself (DRY)

Use a function to replace frequently used code
* Call on the function many times
* automate a function over a list with the <code>apply()</code> family of functions

```{r funciton}
square <- function(x) {
  x^2
}

square(4)

```

## Applying functions (apply/tidyverse)

```{r, eval = F}
day1_mean <- function(x) {
  tibble(id = x$id,
         day1_mean = mean(x$day1, na.rm = T)
  )
}

tki_demo %>%
  split(.$id) %>%
  lapply(day1_mean) %>%
  bind_rows()

```

## Applying functions (apply/tidyverse)

```{r, echo = F}
day1_mean <- function(x) {
  tibble(id = x$id,
         day1_mean = mean(x$day1, na.rm = T)
  )
}

tki_demo %>%
  split(.$id) %>%
  lapply(day1_mean) %>%
  bind_rows()

```